---
title: "Joe Musgrove Pitching Data"
author: "Cory Gargan"
date: "2022-12-08"
output:
  pdf_document: default
  html_document: default
  github_document: default
---
# Taking a Closer Look at Joe Musgrove's Spinrate
#### Did he use a foreign substance to increase spinrate, and if so, did it have a significant impact on his Post Season?  
During the 2022 MLB Playoffs vs. the NY Mets, the Umpires were sent by Mets skipper Buck Showalter to check Pitcher Joe Musgrove for the use of a foreign substance.  
  
The Umpires quick check of the pitcher's ears resulted in a dismissal of the allegation, but was that truly the case?
  
Note: To create my environment I loaded the following libraries:  
  
  * `tidyverse`  
  * `skimr`  
  * `janitor`  
  * `gridExtra`  
  
Along with the dataset:  
  
  * All Data was collected from [MLB Baseball Savant](https://baseballsavant.mlb.com/statcast_search)
  * `MLB_Musgrove-Joe_Reg-Post-Season_Pitch-Data` 
    + This dataset can be found in the github project folder  
    
```{r, message=FALSE, warning=FALSE}
library(tidyverse)  
library(skimr)
library(janitor)
library(gridExtra)
spinrate <- read_csv("~/Documents/Data-Analysis/Github/Portfolio-Projects/01-MLB-Spinrate-Data-in-R/MLB_Musgrove-Joe_Reg-Post-Season_Pitch-Data.csv")
```

### Average Spinrate, Regular vs Post Season  
When looking at the Data there was a clear increase in Post Season Spinrate.  
  
  * 4-Seam Fastball +90.31 RPM  
  * Changeup +101.95 RPM  
  * Curveball +116.26 RPM  
  * Cutter +83.30 RPM  
  * Sinker +79.35 RPM  
  * **Slider +192.60 RPM**  
  
```{r, message=FALSE, warning=FALSE}
# Average Spinrates by Pitch, Regular vs Post Season
avg_spin <- spinrate %>% 
  group_by(season_type, pitch_name) %>% 
  summarize(avg_spinrate = format(round(mean(release_spin_rate), 2), nsmall=2)) %>% 
  arrange(desc(season_type))

# Scatter Plot: Average Spinrate by Pitch, Regular vs Post Season  
ggplot(avg_spin, aes(pitch_name, avg_spinrate, color = season_type)) +
  geom_point(size = 4, shape = 9) +
  labs(title='Average Spinrate', subtitle='Regular Season vs Post Season',
       color='Season Type', x='Pitch Thrown', y='Average Spinrate',
       caption='Data Gathered from baseballsavant.mlb.com') +
  annotate('text', x=5.3, y=11, label='Slider Had the') +
  annotate('text', x=5.3, y=10.6, label='Largest Increase')
```
  
### Average Speed MPH, Regular vs Post Season  
When looking at the Data there was also a increase in Post Season Pitch Speed.  
  
  * 4-Seam Fastball +0.67 MPH  
  * **Changeup +1.58 MPH**  
  * Curveball +0.55 MPH  
  * Cutter +0.71 MPH  
  * Sinker +1.14 MPH  
  * Slider +0.16 MPH  
  
```{r, message=FALSE, warning=FALSE }
# Average Speed MPH by Pitch, Regular vs Post Season
avg_speed <- spinrate %>% 
  group_by(season_type, pitch_name) %>% 
  summarize(average_mph = format(round(mean(effective_speed, na=TRUE), 2), nsmall=2)) %>% 
  arrange(desc(season_type))

# Scatter Plot: Average Speed MPH by Pitch, Regular vs Post Season
ggplot(avg_speed, aes(pitch_name, average_mph, color = season_type)) +
  geom_point(size = 4) +
  labs(title='Average Pitch Speed, MPH', subtitle='Regular Season vs Post Season',
       color='Season Type', x='Pitch Thrown', y='Average Speed, MPH',
       caption='Data Gathered from baseballsavant.mlb.com')
```
  
### With Two Strikes: Percent Thrown for Strikeout, Regular vs Post Season 
The Strikeout percentages decreased for all pitches, **except** the Slider. 
  
  * 4-Seam Fastball -0.02%  
  * Changeup -9.72%  
  * Curveball -5.81%  
  * Cutter -0.71%  
  * Sinker -14.81%  
  * **Slider +2.41%**   
  
```{r, message=FALSE, warning=FALSE}
# With Two Strikes: Percent Thrown for Strikeout by Pitch, Regular vs Post Season
two_strikes <- spinrate %>% 
  group_by(season_type, pitch_name) %>%
  filter(strikes == 2) %>%
  summarize(strikeouts = sum(at_bat_outcome == 'strikeout', na.rm=TRUE),
            times_thrown_2strikes = sum(strikes == 2),
            strikeout_pct = format(round((strikeouts/times_thrown_2strikes)*100, 2), nsmall=2)) %>%
  arrange(desc(season_type)) 

# Column Chart: With Two Strikes: Percent Thrown for Strikeout by Pitch, Regular vs Post Season
ggplot(two_strikes, aes(pitch_name, strikeout_pct, fill = season_type)) +
  geom_col(position = 'dodge') +
  labs(title='With Two Strikes: Percent Thrown for Strikeout', subtitle='Regular Season vs Post Season',
       fill='Season Type', x='Pitch Thrown', y='Strikeout Percentage',
       caption='Data Gathered from baseballsavant.mlb.com') +
  annotate('text', x=5.3, y=11, label='Slider Was The') +
  annotate('text', x=5.3, y=10.6, label='Only Increase')
```
  
### The Slider: Putting it under the Microscope
The Slider was the most improved pitch in regard to increased Spinrate and Strikeout Percentage.  
The Speed also slightly increased, but not what I would consider a significant amount.  
  
  
```{r, message=FALSE, warning=FALSE}
# Slider Strikeout Data
slider_strike <- spinrate %>% 
  group_by(season_type, pitch_name) %>%
  filter(strikes == 2, pitch_name == 'Slider') %>%
  summarize(strikeouts = sum(at_bat_outcome == 'strikeout', na.rm=TRUE),
            times_thrown_2strikes = sum(strikes == 2),
            strikeout_pct = format(round((strikeouts/times_thrown_2strikes)*100, 2), nsmall=2)) %>%
  arrange(desc(season_type))

# Slider Spinrate Data
slider_sr <- spinrate %>% 
  group_by(season_type, pitch_name) %>% 
  filter(pitch_name == 'Slider') %>% 
  summarize(avg_spinrate = format(round(mean(release_spin_rate), 2), nsmall=2)) %>% 
  arrange(desc(season_type))

# Plotting Strikeout & Spinrate Data
plot1 <- ggplot(slider_strike, aes(pitch_name, strikeout_pct, fill = season_type)) +
  geom_col(position = 'dodge') +
  labs(title='Two Strikes: Strikeout Percentage', subtitle='Regular Season vs Post Season',
       fill='Season Type', x='Pitch Thrown', y='Strikeout Percentage',
       caption='Data Gathered from baseballsavant.mlb.com')

plot2 <- ggplot(slider_sr, aes(pitch_name, avg_spinrate, fill = season_type)) +
  geom_col(position = 'dodge') +
  labs(title='Average Spinrate, Slider', subtitle='Regular Season vs Post Season',
       fill='Season Type', x='Pitch Thrown', y='Average Spinrate',
       caption='Data Gathered from baseballsavant.mlb.com')

# Combining the Two Graphs Side-by-Side
grid.arrange(plot1, plot2, ncol=2)
```
  
### In Conclusion:
Although there was never an official investigation into the pitcher's use of a foreign substance during the game, I believe the data tells a different story.  
  
I am of the opinion that the increases in Spinrate and Speed across all pitches points towards the use of a foreign substance on the ball.  
  
In particular with the Slider, the significant Spinrate increase made it a more effective pitch based on the Post Season Strikeout Percentages.    
  
#### We may never find out the truth, but **the data does not lie!**
